\documentclass{article}
\usepackage{authblk}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}

\begin{document}
\title{Collaborative platforms for streamlining workflows in Open Science }

\author[1, 2]{Konrad U. Förstner}
\affil[1]{Institute for Molecular Infection Biology, University of Würzburg, D-97080 Würzburg}
\affil[2]{Research Centre for Infectious Diseases, University of Würzburg, D-97080 Würzburg, Germany}
\author[3]{Gregor Hagedorn}
\affil[3]{Julius Kühn-Institute, Federal Research Center for Cultivated Plants, Berlin, Germany}

\author[4]{Claudia Koltzenburg}
\affil[4]{Managing editor of \href{Cellular Therapy and Transplantation (CTT)}{Cellular Therapy and Transplantation (CTT)}}

\author[5]{M Fabiana Kubke}
\affil[5]{Department of Anatomy with Radiology, University of Auckland}

\author[6]{Daniel Mietchen}
\affil[6]{\href{http://www.science3point0.com/evomri/}{Science 3.0}}

\date{May 9th, 2011}

\maketitle

\begin{abstract}

Despite the internet's dynamic and collaborative nature, scientists
continue to produce grant proposals, lab notebooks, data files,
conclusions etc. that stay in static formats and/or are not published
online and due to this are not accessible to the interested
public. Because of limited adoption of tools that seamlessly integrate
all aspects of a research project (conception, data generation, data
evaluation, peer-reviewing and publishing of conclusions), much effort
is later spent on reproducing or reformatting individual entities
before they can be repurposed as parts of articles or independently.\\

We propose that workflows - both individual and collaborative
performed - could potentially be made more efficient if all steps
within the research cycle would be coherently represented online and
the underlying data were formated, annotated and licensed for
reuse. Such a system would accelerate the process of taking projects
from the conception phase to the publication stage and allow for
continuous updating of the data sets and their interpretation as well
as the integration into other independent projects.\\

Another advantage of such workflows is that the process can be made
transparent, both with respect to the scientific process and to the
contribution of each participant. The latter point is important from a
perspective of motivation, as it enables the allocation of reputation
which creates incentives for scientists to contribute to
projects. Such workflow platforms offering possibilities to fine-tune
the accessibility of their content could gradually pave the path from
the current static mode of research presentation into a more coherent
practice of open science.\\

\end{abstract}


\section{Introduction}

Like most areas of todays life, science has dramatically chanced since
the advent of the internet. However, the transformation that has taken
place until now is just the tip of the iceberg. In the following, we
want to discuss the mostly underutilized potential of representing all
aspect of science in collaboratively used online workflow
platforms. As such platforms could help to realize Open Science, the
practice to grant transparency and access to all data in the research
process, we will shed light on this special aspect and make
recommendations regarding implementations.\\

While there are numerous projects developing and applying so called
Virtual Research Environments (VRE) - also known as
``Collaboratories'' - covering selected stages of the scientific
process, a platform spanning every phase is missing so far
\cite{Carusi}. Technically overcoming such gaps and creating a
seamless transition from bench to publication could speed up the
research and with it the generation, distribution and reuse of
knowledge.

\section{The scientific workflow in open VREs}

\subsection{Conception and project planning}

Independent of the nature of a research endeavor - hypotheses-driven
or data-driven, performed by a single person or a team - a solid
conception phase is the crucial basis for each project. Despite
today's common practice, according to which this is done by a small
group of people, harvesting collective intelligence could help to
avoid redundant research and to improve instead the design of the
study. As the complexity and scope of scientific projects are
increasing, the application of project management tools can be useful
for managing the processes and parties involved.

\subsection{Experiments and data generation}

Today, data generation in academic research relies comparatively
strongly on manual labor. While this is mostly due to the low price of
labor force resulting from the academic system and the limited
interdisciplinary eduction of science and engineering, the potential
of automation is extremely high but mostly neglected. Not only could
the efficiency of invested labor be improved, but also reproducibility
could be significantly increased. To make this affordable for the
broader research community, a shift from siloed proprietary devices to
well-documented pieces of standardized, open-source hardware developed
by the scientific community itself is needed and could take place in
cooperation with potential vendors. Open hardware platforms like
Arduino could be starting points for such a development. The devices
could and should enrich the primary data with further metadata,
convert them into semantified formats and directly upload the output
into online repositories.\\

One promising example which visualizes the potential of such
automation of usually quite labor-intensive research is the robot
scientist ADAM \cite{Ross}. The streamlining of mechanical steps and
the evaluation of results would benefit from formal languages that
describe the necessary procedures and make the design and exchange of
experimental setups easy \cite{Soldatova}. As a long term goal, scientists
would mostly engage in programming experiments and engineering the
system to automate those steps that have been performed manually so
far. The motto ``work on the system, not in the system'' should guide
this development.

\subsection{Data release}

The online release of experimentally generated data should be done
shortly after the generation and can happen potentially in real
time. Downstream analysis inside of the research project but also the
reuse by other parties should be kept in mind when selecting data
formats. These should, as far as possible, be non-proprietary, machine
readable (semantically enriched) and common for the respective domain
of research. If no format fulfills all these requirements, the
conversion into alternative formats should be permitted. Access to the
data could take place via a web interface or domain specific
clients. Especially for large or highly accessed data sets, the
additional distribution via peer-to-peer networks is recommended.

\subsection{Data analysis}

Since every step in the data analysis should be transparent and easily
reproducible, it should take place preferably in the proposed
platform, too. Systems like the analysis workflow tool Taverna
\cite{Hull} could be used for such processing. Already today, many
research institution offer grid computing infrastructure for such
purposes. Analyses using external tools, especially GUI-tools that do
not offer any possibility to log the performed actions, should be
avoided if possible, as otherwise documentation has to be created
manually. A potential side-effect of running computationally intensive
analyses on shared systems is a more economical usage of the needed
infrastructure. As done for the raw experimental data, the protocols
and the result of the data processing should be documented and stored
in repositories to be accessible.

\subsection{Knowledge generation}

The result of the processing as well as the raw data can be used by
scientists - or machines \cite{Schmidt} - to draw conclusions and to generate
knowledge out of the available information in a well documented
way. The platform should assist to make this happen collaboratively by
offering commenting and rating of statements. Discussions - text,
audio- and/or video-based - should be recorded to make the path to
finding reconstructible.

\subsection{Final publication}

As documentation of every step is an inherent feature of the workflow,
the final publications resulting from a study can be short reports
linking to the major outcomes and putting them into the scientific
context. The platform should offer functionalities to perform open
peer-review of this final report.

\section{Implementation}

\subsection{Technology}

As shown above, the many building blocks of a complete scientific
workflow already exist and only need to be connected seamlessly. The
development of open standards defining the required interfaces of
these parts could enable different parties to assemble the pieces into
a consistent workflow and to add further needed parts. This would
offer the possibility to implement a platform either as one monolithic
application or as separate interacting and exchangeable units.

\subsection{Funding}

Of similar importance as the technical realization is the adaptation
of scientific culture and funding policies. While research
institutions like the National Institutes of Health (US) or the
Welcome Trust (UK) already require open access for final peer-review
manuscripts that results from research they funded \cite{NIH, Welcome}, the
regulations are much weaker for the underlying data, and almost
nonexistent for proper annotation.

\subsection{Licensing}

As the default copyright restrictions in most jurisdictions hamper the
reuse of data, it is highly desirable that, with very few exceptions,
each entity generated in the research process is explicitly published
under a less restrictive license e.g. the ones offered by Creative
Commons \cite{CC} or is released into the public domain (as the public
domain is not a concept in every country CC0 \cite{CC0} could be applied
equivalently).

\subsection{Reputation}

The gain of reputation is the most important incentive for
scientists. It is currently mostly determined on the basis of
publications in scientific journals and the related measure of success
in funding applications. As every contribution to a research project
can be attributed to a distinct person and could be rated by others,
the allocation of reputation is an inherent element of the proposed
platform. The connection to research identifiers like ORCID \cite{ORCID} and
the analysis of such microcontributions could assemble a precise image
of a scientist's skills and achievements.


\section{Challenges}

As stated above, considering the allocation of reputation and funding
in science is crucial when redesigning scientific processes. To bridge
a transient phase until the suggested political changes have taken
place, fine granular access control in the research workflow platform
could permit that the technology is adapted by scientists despite
objection regarding the loss of reputation. With such a control in
place, the full process could be opened up after the final publication
or at any other desired time.\\

It is very unlikely that there will be one single platform that can
fulfill the requirements of each scientific domain. Instead, for every
community, a platform of this kind might be based on a more general
solution.

\begin{thebibliography}{9}

\bibitem{Carusi} Annamaria Carusi, Torsten Reimer. Virtual Research
  Environment - Landscape Collaborative Study. {\em JISC}. pp. 72-24
  2010.

\bibitem{Ross} Ross D. King, Jem Rowland, Stephen G. Oliver, Michael
  Young, Wayne Aubrey, Emma Byrne, Maria Liakata, Magdalena Markham,
  Pinar Pir, Larisa N. Soldatova, Andrew Sparkes, Kenneth E. Whelan,
  Amanda Clare. The automation of science. {\em Science}. 3 April 2009:
  Vol. 324 no. 5923 pp. 85-89

\bibitem{Soldatova} Larisa N Soldatova, Ross D King. An ontology of
  scientific experiments. {\em J R Soc Interface}. 2006 December 22;
  3(11): 795–803.

\bibitem{Hull} Duncan Hull, Katy Wolstencroft, Robert Stevens, Carole
  Goble, Mathew R. Pocock, Peter Li, Tom Oinn.  Taverna: a tool for
  building and running workflows of services {\em Nucleic Acids
    Res}. 2006 Jul 1;34(Web Server issue):W729-32.

\bibitem{Schmidt} Michael Schmidt and Hod Lipson. Distilling Free-Form
  Natural Laws from Experimental Data. {\em Science}. 3 April 2009:
  Vol. 324 no. 5923 pp. 81-85

\bibitem{NIH} \url{http://publicaccess.nih.gov/policy.htm}

\bibitem{Welcome}
  \url{http://www.wellcome.ac.uk/About-us/Policy/Policy-and-position-statements/WTD002766.htm}

\bibitem{CC} \url{http://creativecommons.org/}

\bibitem{CC0} \url{http://creativecommons.org/publicdomain/zero/1.0/}

\bibitem{ORDID} \url{http://orcid.org/}

\end{thebibliography}

\end{document}
